---
title: "FML implementation performance evaluation"
author: "Insert name here"
date: "4/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include=FALSE}
load_or_install <- function(package) {
    if (!require(package, character.only = TRUE)) {
        install.packages(package)
        if (!require(package, character.only = TRUE)) {
            stop(paste0("Cannot install package ", package));
        }
    }
}

load_or_install("dplyr")
load_or_install("readr")
load_or_install("tidyr")
load_or_install("ggplot2")
load_or_install("scales")
load_or_install("grid")
load_or_install("gridExtra")
load_or_install("stringr")
load_or_install("tools")

draw_grid <- function(sets, f) {
  graphs <- lapply(sets, FUN=f)
  #print(graphs)
  #remove_label <- function(g) { g + theme(legend.position = "none") }
  #graphs <- lapply(graphs, FUN=remove_label)
  do.call(grid.arrange, graphs)
}
```

```{r data, include=FALSE}
timing_log <- read_csv("timing_log.csv")
#timing_log <- read_csv("~/Workspace/FMLBench/timing_log.csv")
```
This report describes the results of a performance comparison between multiple implementations of the FML language using a suite of benchmark programs.

## System description

The benchmark was executed using the followins system specs.

| Parameter                      | Spec                  | Notes |
| :---------------               | :-------------        | :---- |
| Processor                      | TODO                  |       |
| Memory                         | TODO                  |       |
| OS & kernel                    | TODO                  |       |

## Benchmarks

The experiment was run using three benchmark programs. Each benchmark was executed 10 times (iterations 1-10).

| Label                          | Description           | URL                                                                       |
| :---------------               | :-------------        | :---------                                                                |
| `nested_loops`                 | Nested loops          | https://github.com/kondziu/FMLBench/blob/main/benchmarks/nested_loops.fml |
| `langtons_ant`                 | Langton's ant         | https://github.com/kondziu/FMLBench/blob/main/benchmarks/langtons_ant.fml |
| `brainfuck`                    | Brainfuck interpreter | https://github.com/kondziu/FMLBench/blob/main/benchmarks/brainfuck.fml    |
| `sudoku`                       | Sudoku solver         | https://github.com/kondziu/FMLBench/blob/main/benchmarks/sudoku.fml       |

### Nested loops

A very plain and simple benchmark - just a few nested loops doing nothing. It's meant to test allocation of integers and methods operating on them (`+`, `<`). Because of its simplicity a lot of it could be optimized by a JIT. Though the fact that it consists of a single (entry point) function, might make it problematic for the simplest method JITs.

### Langton's ant

An ant moves on a world of square tiles. The ant has position on a specific tile and an orientation (up, down, left, right). The tiles are initiall all white, but each tile can be either black or white. The program runs for 201 steps. In each step the ant moves one tile according to the following rules.

* If the ant is on a while tile, it changes the tile's color to black, rotates clockwise and moves one tile forward.
* If the ant is on a black tile, it changes the tile's color to white, rotates counterclockwise and moves one tile forward.

The board and the ant are printed after each step. The world starts empty and expands whenever the ant moves to a new tile.

[[wiki]](https://en.wikipedia.org/wiki/Langton%27s_ant)

### Brainfuck interpreter

An implementation of the Brainfuck language executing a specific program.

The language contains 8 instructions `.,<>[]+-`, 7 of which are expressable in FML (FML cannot read from stdin, preventing it from implementing `,`). The interpreter has a memory of 512 integers and a memory pointer. The interpreter moves the pointer to the left and right (`<` and `>`), increments and decrements the values at the pointer (`+` and `-`), and prints out the value of a cell as ASCII (`.`). `[` and `]` imeplement loops: `[` checks if the value of the cell at the pointer is 0, and proceeds to a matching `]` if it is or executes the next statement if it isn't. `]` returns to the matching `[`.

In this implementation the memory consists of 512 32bit integers. The memory pointer is initially positioned in the middle.

The program executed by the benchmark prints out a poem in 429 instructions.

[[wiki]](https://en.wikipedia.org/wiki/Brainfuck)

### Sudoku solver

A brute force sudoku solver. It starts from the top-left corner of the board, adds a number from 1-9 (in order) in the first available empty spot and checks whether the board is still valid. It continues until the end of the board and backtracks to if the board becomes invalid at any point.

[[wiki]](https://en.wikipedia.org/wiki/Sudoku)

## Implementations

The experiment was run using the following implementations of the FML language. 

| Label                          | Description                       | URL                                            |
| :---------------               | :-------------                    | :---------                                     |
| `/cfml/fml`                    | The reference FML implementation. | https://gitlab.fit.cvut.cz/vlasami6/cfml       |
| TODO                           | TODO                              |                                                |
| TODO                           | TODO                              |                                                |

The `/cfml/fml` implementation is written in C The stack consists of a bytecode compiler and a bytecode interpreter. Execution times include parsing, compilation, and execution. The implementation does not garbage collect, so it does not incur garbage-collection-related overheads.

TODO

## Correctness 

The benchmarks were sanity checked after execution by comparing outputs.

```{r correctness, echo=FALSE, fig.width=15}
correctness1 <- setNames(c("#bd320b","#21e6ff"), c(FALSE, TRUE))
correctness2 <- setNames(c("red","blue"), c(FALSE, TRUE))
correctness_for_implementation <- function(implementation, rows=1) {
  timing_log %>% 
    filter(`FML implementation` == !!implementation)  %>%
    ggplot(aes(x=iteration, y=benchmark)) + 
    geom_tile(aes(fill=correct, color=correct, width=0.7, height=0.7), linewidth=0.5) + 
    scale_x_discrete(breaks=1:10, limits=factor(1:10)) +
    scale_fill_manual(values=correctness1) +
    scale_color_manual(values=correctness2) +
    #scale_fill_continuous(function(x) {print(x); ifelse(x < 0, "#bd320b","#21e6ff")}) +
    theme_dark() + 
    theme(panel.grid.major = element_blank(), axis.ticks = element_blank(), axis.title.y = element_blank(), legend.position = "none") +
    ggtitle(implementation)
}
implementations <- timing_log %>% pull(`FML implementation`) %>% unique()
draw_grid(implementations, correctness_for_implementation)
```

#### Observations

TODO 

## Execution time

```{r execution-time, echo=FALSE, fig.width=15}
violins_for_benchmark <- function (benchmark, title) {
  ggplot(timing_log %>% filter(benchmark==!!benchmark), aes(x=`FML implementation`, y=millis)) + 
    geom_violin(trim=FALSE, aes(color=`FML implementation`, fill=`FML implementation`)) + 
    theme_dark() + 
    theme(panel.grid.major = element_blank(), axis.ticks.x = element_blank(), legend.position = "none", axis.title.x = element_blank()) +
    ylab("Elapsed time [ms]") +
    ggtitle(title)
}

grid.arrange(
  violins_for_benchmark("nested_loops", "Nested loops") +  theme(axis.title.y = element_blank()),
  violins_for_benchmark("langtons_ant", "Langton's ant"),
  violins_for_benchmark("brainfuck", "Brainfuck interpreter") +  theme(axis.title.y = element_blank()),
  violins_for_benchmark("sudoku", "Sudoku solver") +  theme(axis.title.y = element_blank()),
  ncol=2)
```

#### Observations

TODO 

## Memory usage

(These can take a while to generate).
```{r memory_prep, echo=FALSE, message=FALSE, results='hide'}

area_for_single_execution <- function(log_file, show_gc_time=FALSE) {
  matches <- str_match(log_file, regex("heap_log/([^:]*):([^:]*).csv"))
  benchmark <- str_replace_all(matches[2], "\\\\", "/")
  implementation <- str_replace_all(matches[3], "\\\\", "/")

  #cat(paste0("Preparing graph for: ", benchmark , " + ", implementation, "\n"));
  #browser()

  heap_log <- read_csv(log_file, col_types=cols()) %>% mutate(elapsed=timestamp - min(timestamp)) 
  #%>% mutate(gc = ifelse(event=="B", timestamp, as.integer(NA)))
  max_heap_size <- heap_log %>% pull(heap) %>% max
  #heap_log <- heap_log %>% mutate(heap = ifelse(event == "S", 0, heap))
                                            
  graph <- ggplot (heap_log, aes(x=elapsed, y=heap))
  #geom_rect(xmin = -Inf, xmax = 0,   ymin = -Inf, ymax = 0,   fill = "red")

  if (show_gc_time) {
    gc_begin <- heap_log %>% filter(event == "B") %>% 
      rename(start_timestamp=timestamp, start_event=event, start_heap=heap, start_elapsed=elapsed)
    gc_end <- heap_log %>% filter(event == "A") %>% 
      rename(end_timestamp=timestamp, end_event=event, end_heap=heap, end_elapsed=elapsed)
    if (gc_begin %>% count %>% pull(n) != gc_end %>% count %>% pull(n)) {
      warning(paste0(
        "The number of GC start events (B) is different than GC end events (A): ",
        gc_begin %>% count %>% pull(n), "!=", gc_end %>% count %>% pull(n)
      ))
    }
    gc <- cbind(gc_begin, gc_end)
    graph <- graph + geom_rect(data = gc, inherit.aes = FALSE,
      aes(xmin = start_elapsed, xmax = end_elapsed, ymin = -Inf, ymax = +Inf),
      color = "white")
  }
  
  graph <- graph + geom_area() #+ geom_vline(aes(xintercept=gc))
  # if (show_max) {
  #   graph <- graph + geom_hline(yintercept=max_heap_size, linetype="dashed", color = "white")
  # }
  
  graph +
    theme_dark() + 
    theme(panel.grid.major = element_blank(), axis.ticks.x = element_blank(), legend.position = "none") +
    scale_y_continuous(labels=function(x)
      ifelse(x >= 1024 * 1024 * 1014, paste0(as.integer(x / 1024 / 1024 / 1024),"GB"),
             ifelse(x >= 1024 * 1024, paste0(as.integer(x / 1024 / 1024), "MB"),
                    ifelse(x >= 1024, paste0(as.integer(x / 1024), "KB"),
                           paste0(x, "B"))))) +
    scale_x_continuous(labels=function(x)
      ifelse(x >= 60 * 1000 * 1000 * 1000, paste0(as.integer(x / 60 / 1000 / 1000 / 1000),"min"),
          ifelse(x >= 1000 * 1000 * 1000, paste0(as.integer(x / 1000 / 1000 / 1000),"s"),
                 ifelse(x >= 1000 * 1000, paste0(as.integer(x / 1000 / 1000), "ms"),
                        ifelse(x >= 1000, paste0(as.integer(x / 1000), "μs"),
                               paste0(x, "ns")))))) +
    xlab("Elapsed time") +
    ylab("Heap size") +
    ggtitle(paste0("Memory usage in ", benchmark)) + 
    labs(subtitle=implementation)
}

plots <- list()
all_files <- list.files(recursive=TRUE)
files <- all_files[grep(all_files, pattern="^heap_log:")]
for (file in files) {
  plots[[length(plots) + 1]] <- area_for_single_execution(file, show_gc_time=TRUE) # set show_gc_time to TRUE to show (very slim) white rectangles indicating time spent inside GC, or FALSE to get rid of them
}
```

```{r memory, echo=FALSE, fig.width=15, fig.height=20}
if (length(plots) > 0) {
  do.call(grid.arrange, plots)
}
```

#### Observations

TODO
